---
title: "LVIS Analysis Autism Mansucript 2022"
author: "Elena Tenenbaum and Adam Naples"
date: "3/8/2021"
output: html_document
---
```{r}
#install.packages('psych')
library(expss)
library(psych)
library(ggplot2)
library(ggpubr)
library(ggsci)
library(tidyverse)
library(tidyr)
library(broom)
library(corrplot)
library(mice)
library(paran)
library(Hmisc)
library(ppcor)
library(gtsummary)
library(readxl)
```

```{r}
#read in data file
set.seed(2747)
lvisCollapse <- read.csv("LVIS_Data.csv")

#remove out of age range participants
lvisCollapse<-lvisCollapse[which(lvisCollapse$age<10),]

#extracts LVIS data only
lvis_data<-lvisCollapse[,20:53]
#imputes missing using MICE
imputed_Data <- mice(lvis_data, m=5, maxit = 50, method = 'pmm', seed = 500)
summary(imputed_Data)
#randomly selects 2nd imputed dataset
completeData <- complete(imputed_Data,2)

#runs Monte Carlo simulation to determine number of factors to extract - returns 6 
paran(completeData, iterations=0, centile=0, quietly=FALSE,
status=TRUE, all=FALSE, cfa=FALSE, graph=FALSE,
color=TRUE, col=c("black","red","blue"),
lty=c(1,2,3), lwd=1, legend=TRUE, file="",
width=640, height=640, grdevice="png", seed=0, mat=NA, n=NA)

cortest.bartlett(completeData)
KMO(completeData)

#this is tough to interpret and was rejected in favor of 2 factor model below
pc6<-principal(completeData,nfactors=6,rotate="oblimin")
pc6
print.psych(pc6,cut=.3,sort=TRUE)
plot(pc6$values,type="b")

#this makes sense given scree plot and is more interpretable
pc2<-principal(completeData,nfactors=2,rotate="oblimin")
pc2
print.psych(pc2,cut=.3,sort=TRUE)
plot(pc2$values,type="b")
```

#creates LVIS score based on the first factor of 2 factor PCA
```{r}
#lvis_data<-completeData #commenting this line so that data will be consistent and stop pulling from imputed data set. this may result in errors later due to missing data.

#replacing NAs in lvis_data with 0 so that those scores will be included in full sample
lvis_data[is.na(lvis_data)]=0

lvis_data$lvis_SS<-(lvis_data$lvis_7+lvis_data$lvis_28+lvis_data$r3+lvis_data$lvis_11+lvis_data$lvis_31+lvis_data$lvis_33+lvis_data$lvis_26+lvis_data$lvis_32+lvis_data$lvis_27+lvis_data$lvis_30+lvis_data$lvis_35+lvis_data$r2+lvis_data$lvis_15+lvis_data$lvis_34+lvis_data$r1+lvis_data$lvis_14+lvis_data$lvis_16)
lvisCollapse$lvis_SS<-lvis_data$lvis_SS
summary(lvis_data$lvis_SS)
sd(lvis_data$lvis_SS)


#creates an atypical score based on factor 2 of PCA
lvis_data$lvis_AT<-(lvis_data$lvis_23+lvis_data$lvis_24+lvis_data$lvis_19+lvis_data$lvis_37+lvis_data$lvis_21+lvis_data$lvis_10+lvis_data$lvis_18+lvis_data$lvis_36+lvis_data$lvis_22+lvis_data$lvis_20+lvis_data$lvis_38+lvis_data$lvis_12+lvis_data$lvis_13+lvis_data$lvis_25)
lvisCollapse$lvis_AT<-lvis_data$lvis_AT
summary(lvis_data$lvis_AT)
sd(lvis_data$lvis_AT)
```
#CORRELATION MATRIX
```{r}
#provides correlations for the factors included in the lvis_SS, replacing redudant items with the combined items.
L_VIS_Corr_PCA<-lvis_data[,c("lvis_7",
"lvis_28", "r3","lvis_11","lvis_31","lvis_33","lvis_26","lvis_32","lvis_27","lvis_30","lvis_35","r2","lvis_15",
"lvis_34","r1","lvis_14","lvis_16","lvis_29","lvis_8","lvis_23","lvis_24","lvis_19","lvis_37","lvis_21",
"lvis_10","lvis_18","lvis_36","lvis_22","lvis_20","lvis_38","lvis_13","lvis_12","lvis_25","lvis_9")]


#prints correlation table
mcor<-round(cor(L_VIS_Corr_PCA,method="pearson",use="complete.obs"),2)
mcor[mcor == 1] <- NA
summary(mcor)

max(mcor)

#identifies correlations greater than .8
mcor[mcor>.8&mcor<1] #nothing greater than .8 identified
```


Demographics
```{r}
lvisCollapse %>% 
  select(age,sex,ados_module,group3) %>% 
  tbl_summary(by = group3,statistic = all_continuous() ~ "{mean} ({sd})") %>% 
  add_overall() %>% 
  add_n() %>% 
  add_stat_label()

#Table 2
lvisCollapse %>% 
  select(group3,lvis_SS,pls_acss,pls_ecss,pls_totae,vine_com_std_score,vine_soc_std_score,vine_dls_std_score,vine_ms_std_score,binet_nv_stand_score,binet_v_stand_score) %>%
  tbl_summary(by = group3,statistic = list(all_continuous() ~ "{mean} ({sd})"), digits = list(all_continuous() ~ c(1, 1))) %>% 
  add_n() %>% 
  add_overall() %>%
  add_stat_label() 
```




VABS scatterplots
```{r}
#runs correlations/scatterplots with sumscore for VABS

#assigns labels for groups
lvisCollapse$group3str<-car::recode(lvisCollapse$group3,"1='ASD';2='DLD';3='TD'")

#pulls together VABS scores and lvis_SS with age and group
L_VIS_VABS <- lvisCollapse[,c("vine_com_std_score","vine_dls_std_score","vine_soc_std_score", "vine_ms_std_score","lvis_SS","group3str","age")]
#creates long file so that PLS scores can be facet wrapped in figures
VABS<-gather(L_VIS_VABS,key="VABS", value="Score", na.rm=FALSE,-lvis_SS,-group3str,-age)
#provides labels for the VABS scores
VABS_labels <- c(vine_com_std_score="Communication",vine_dls_std_score="Daily Living",vine_soc_std_score="Socialization",vine_ms_std_score="Motor")

#scatterplot for VABS with color
Figure_2c<-ggplot(VABS, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  labs(title="",
       x="LVIS Composite Communication Scores", y = "Vineland Standard Score")   +
  facet_grid(. ~ VABS)+
  facet_grid(. ~ VABS, labeller=labeller(VABS = VABS_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank())
Figure_2c


#scatterplot for VABS with grayscale
Figure_2g<-ggplot(VABS, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  scale_color_manual(values=c("#000000", "#949494", "#C6C6C6"))+
  labs(title="",
       x="LVIS Composite Communication Scores", y = "Vineland Standard Score")   +
  facet_grid(. ~ VABS)+
  facet_grid(. ~ VABS, labeller=labeller(VABS = VABS_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank())

#creates scatter plots for VABS scores with blues for poster
Figure_2b<-ggplot(VABS, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(size = 1.5,aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
scale_color_manual(values=c("blue4", "#56B4E9", "#0072B2"))+
labs(title="",
       x="\n LVIS Composite Communication Scores", y = "Vineland Standard Score \n")   +
  facet_grid(. ~ VABS)+
  facet_grid(. ~ VABS, labeller=labeller(VABS = VABS_labels))+
  scale_fill_discrete(name="Group")+
theme(legend.title=element_blank(), text=element_text(size=19,  family="serif"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.text = element_text(margin = margin(l = 10, unit = "pt")))+
theme(axis.text.x= element_text(size=10))+
theme(axis.text.y= element_text(size=10))
Figure_2b


tiff(filename = "Figure_2.tif", height = 12, width = 17, units = 'cm',compression = "lzw", res = 300)
plot(Figure_2c)
dev.off()

```
#SCATTERPLOTS FOR PLS SCORES
#this is still missing the grayscale and blue color schemes - waiting for the standard scores of PLS and will then finalize. 
```{r}
L_VIS_PLS <- lvisCollapse[,c("pls_totss","pls_acss","pls_ecss","lvis_SS","group3str","age")]
PLS<-gather(L_VIS_PLS,key="PLS", value="Score", na.rm=FALSE,-lvis_SS,-group3str,-age)
PLS_labels <- c(pls_totss="Total",pls_acss="Receptive",pls_ecss="Expressive")

#scatterplot for PLS with color
Figure_1c<-ggplot(PLS, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  labs(title="",
       x="LVIS Composite Communication Scores", y = "PLS Standard Score")   +
  facet_grid(. ~ PLS)+
  facet_grid(. ~ PLS, labeller=labeller(PLS = PLS_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank())
Figure_1c


tiff(filename = "Figure_1.tif", height = 12, width = 17, units = 'cm',compression = "lzw", res = 300)
plot(Figure_1c)
dev.off()


```
#SCATTERPLOTS FOR SB SCORES
```{r}
L_VIS_SB <- lvisCollapse[,c("binet_v_stand_score","binet_nv_stand_score","lvis_SS","group3str","age")]
SB<-gather(L_VIS_SB,key="SB", value="Score", na.rm=FALSE,-lvis_SS,-group3str,-age)
SB_labels <- c(binet_v_stand_score="Verbal",binet_nv_stand_score="NonVerbal")

#scatterplot for SB with color
Figure_3c<-ggplot(SB, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  labs(title="",
       x="LVIS Composite Communication Scores", y = "Stanford Binet Standard Score")   +
  facet_grid(. ~ SB)+
  facet_grid(. ~ SB, labeller=labeller(SB = SB_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank())
Figure_3c

#scatterplot for SB with grayscale
Figure_3g<-ggplot(SB, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  scale_color_manual(values=c("#000000", "#949494", "#C6C6C6"))+
  labs(title="",
       x="LVIS Composite Communication Scores", y = "Stanford Binet Score")   +
  facet_grid(. ~ SB)+
  facet_grid(. ~ SB, labeller=labeller(SB = SB_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank())
Figure_3g

#scatterplot for SB with blues for poster
Figure_3b<-ggplot(SB, aes(x=lvis_SS, y=Score, shape=group3str, color=group3str)) +
  theme_set(theme_minimal())+
  geom_point(aes(shape=group3str)) +    # Use hollow circles
  geom_smooth(method=lm)+   # Add linear regression line (by default includes 95% confidence region)
  scale_color_manual(values=c("blue4", "#56B4E9", "#0072B2"))+
  labs(title="",
       x="LVIS Composite Communication Scores", y = "Stanford Binet Score")   +
  facet_grid(. ~ SB)+
  facet_grid(. ~ SB, labeller=labeller(SB = SB_labels))+
  scale_fill_discrete(name="Group")+
  theme(legend.title=element_blank(), text=element_text(size=19,  family="serif"),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),legend.text = element_text(margin = margin(l = 10, unit = "pt")))+
theme(axis.text.x= element_text(size=10))+
theme(axis.text.y= element_text(size=10))
Figure_3b


tiff(filename = "Figure_3.tif", height = 12, width = 17, units = 'cm',compression = "lzw", res = 300)
plot(Figure_3c)
dev.off()


```
Correlation Matrices all participants
```{r}


flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

#Correlations with full sample
L_VIS_VABS_Corr <- L_VIS_VABS[,c("vine_com_std_score","vine_dls_std_score","vine_soc_std_score", "vine_ms_std_score","lvis_SS","age")]
VABS_CM<-rcorr(as.matrix(L_VIS_VABS_Corr))
all_VABS<-tidy(VABS_CM)
all_VABS

L_VIS_PLS_Corr <- L_VIS_PLS[,c("pls_totss","pls_acss","pls_ecss","lvis_SS","age")]
PLS_CM<-rcorr(as.matrix(L_VIS_PLS_Corr))
all_PLS<-tidy(PLS_CM)
all_PLS

L_VIS_SB_Corr <- L_VIS_SB[,c("binet_v_stand_score","binet_nv_stand_score","lvis_SS","age")]
SB_CM<-rcorr(as.matrix(L_VIS_SB_Corr))
all_SB<-tidy(SB_CM)
all_SB
```
Correlation Matrices ASD only
```{r}
#Correlations with ASD Only
L_VIS_VABS_Corr_ASD <- L_VIS_VABS[which(L_VIS_VABS$group3str=="ASD"),c("vine_com_std_score","vine_dls_std_score","vine_soc_std_score", "vine_ms_std_score","lvis_SS","age")]
VABS_CM_ASD<-rcorr(as.matrix(L_VIS_VABS_Corr_ASD))
td_VABS_ASD<-tidy(VABS_CM_ASD)
td_VABS_ASD

L_VIS_PLS_Corr_ASD <- L_VIS_PLS[which(L_VIS_PLS$group3str=="ASD"),c("pls_totss","pls_acss","pls_ecss","lvis_SS","age")]
PLS_CM_ASD<-rcorr(as.matrix(L_VIS_PLS_Corr_ASD))
td_PLS_ASD<-tidy(PLS_CM_ASD)
td_PLS_ASD

L_VIS_SB_Corr_ASD <- L_VIS_SB[which(L_VIS_SB$group3str=="ASD"),c("binet_v_stand_score","binet_nv_stand_score","lvis_SS","age")]
SB_CM_ASD<-rcorr(as.matrix(L_VIS_SB_Corr_ASD))
td_SB_ASD<-tidy(SB_CM_ASD)
td_SB_ASD
```

```{r}

#Correlations with DLD Only
L_VIS_VABS_Corr_DLD <- L_VIS_VABS[which(L_VIS_VABS$group3str=="DLD"),c("vine_com_std_score","vine_dls_std_score","vine_soc_std_score", "vine_ms_std_score","lvis_SS","age")]
VABS_CM_DLD<-rcorr(as.matrix(L_VIS_VABS_Corr_DLD))
td_VABS_DLD<-tidy(VABS_CM_DLD)
td_VABS_DLD

L_VIS_PLS_Corr_DLD <- L_VIS_PLS[which(L_VIS_PLS$group3str=="DLD"),c("pls_totss","pls_acss","pls_ecss","lvis_SS","age")]
PLS_CM_DLD<-rcorr(as.matrix(L_VIS_PLS_Corr_DLD))
td_PLS_DLD<-tidy(PLS_CM_DLD)
td_PLS_DLD

L_VIS_SB_Corr_DLD <- L_VIS_SB[which(L_VIS_SB$group3str=="DLD"),c("binet_v_stand_score","binet_nv_stand_score","lvis_SS","age")]
SB_CM_DLD<-rcorr(as.matrix(L_VIS_SB_Corr_DLD))
td_SB_DLD<-tidy(SB_CM_DLD)
td_SB_DLD

```

```{r}
#Correlations with TD Only
L_VIS_VABS_Corr_TD <- L_VIS_VABS[which(L_VIS_VABS$group3str=="TD"),c("vine_com_std_score","vine_dls_std_score","vine_soc_std_score", "vine_ms_std_score","lvis_SS","age")]
VABS_CM_TD<-rcorr(as.matrix(L_VIS_VABS_Corr_TD))
td_VABS_TD<-tidy(VABS_CM_TD)
td_VABS_TD

L_VIS_PLS_Corr_TD <- L_VIS_PLS[which(L_VIS_PLS$group3str=="TD"),c("pls_totss","pls_acss","pls_ecss","lvis_SS","age")]
PLS_CM_TD<-rcorr(as.matrix(L_VIS_PLS_Corr_TD))
td_PLS_TD<-tidy(PLS_CM_TD)
td_PLS_TD

L_VIS_SB_Corr_TD <- L_VIS_SB[which(L_VIS_SB$group3str=="TD"),c("binet_v_stand_score","binet_nv_stand_score","lvis_SS","age")]
SB_CM_TD<-rcorr(as.matrix(L_VIS_SB_Corr_TD))
td_SB_TD<-tidy(SB_CM_TD)
td_SB_TD

```

ADOS correlations
```{r}
#pull in ADOS scores 

L_VIS_DATA_ADOS <- read.csv("LVIS_Data.csv")
#reduces to appropriate age range
L_VIS_DATA_ADOS <-L_VIS_DATA_ADOS[which(L_VIS_DATA_ADOS$age<10),]

#merge lvisCollapse with ADOS CSS
lvisCollapse <- merge(lvisCollapse,L_VIS_DATA_ADOS,by="ID")
#this includes data error repair identified by Adam in email exchange dated 1/19/22

L_VIS_ADOS <- lvisCollapse[,c("ADOS_CSS.y","ADOS_CSS_SA.y","ADOS_CSS_RRB.y","lvis_SS","lvis_AT","group3str","age.y")]
#creates long file so that ADOS scores can be facet wrapped in figures
ADOS<-gather(L_VIS_ADOS,key="ADOS", value="Score", na.rm=FALSE,-lvis_SS,-group3str,-age.y)
#provides labels for the ADOS scores
ADOS_labels <- c(ADOS_CSS.y="ADOS Severity Score",ADOS_CSS_SA.y="ADOS Social Affect Severity Score",ADOS_CSS_RRB.y="ADOS RRB Severity Score")

L_VIS_ADOS$ADOS_CSS.y<-as.double(L_VIS_ADOS$ADOS_CSS.y)
L_VIS_ADOS$ADOS_CSS_SA.y<-as.double(L_VIS_ADOS$ADOS_CSS_SA.y)
L_VIS_ADOS$ADOS_CSS_RRB.y<-as.double(L_VIS_ADOS$ADOS_CSS_RRB.y)
L_VIS_ADOS$age.y<-as.double(L_VIS_ADOS$age.y)

#ASD only
L_VIS_ADOS_Corr_ASD <- L_VIS_ADOS[which(L_VIS_ADOS$group3str=="ASD"),c("ADOS_CSS.y","ADOS_CSS_SA.y","ADOS_CSS_RRB.y","lvis_SS","lvis_AT","age.y")]
ADOS_CM_ASD<-rcorr(as.matrix(L_VIS_ADOS_Corr_ASD))
td_ADOS_ASD<-tidy(ADOS_CM_ASD)
td_ADOS_ASD



```
Calculate LVIS Communication Ability Scores by Group and MV status
```{r}

lvis_MV<-read.csv("LVIS_MV_NID.csv")
lvisCollapse <- merge(lvisCollapse,lvis_MV,by="ID")
MV_ASD<- lvisCollapse[which(lvisCollapse$group3str=="ASD"),]

MV_ASD %>% 
  select(age.x,sex.x,ados_module.x,group3.y,lvis_NVnow) %>% 
  tbl_summary(by = lvis_NVnow,statistic = all_continuous() ~ "{mean} ({sd})") %>% 
  add_n() %>% 
  add_stat_label()

MV_ASD %>% 
  select(group3.y,lvis_SS,pls_acss.x,pls_ecss.x,pls_totae.x,vine_com_std_score.x,vine_soc_std_score.x,vine_dls_std_score.x,vine_ms_std_score.x,binet_nv_stand_score.x,binet_v_stand_score.x,lvis_NVnow, ADOS_CSS.x, ADOS_CSS_SA.x,ADOS_CSS_RRB.x) %>%
  tbl_summary(by = lvis_NVnow,statistic = list(all_continuous() ~ "{mean} ({sd})"), digits = list(all_continuous() ~ c(1, 1))) %>% 
  add_n() %>% 
  add_stat_label() 

#it won't give means of ADOS SS cause it thinks they are factors, so kludgy solution to follow

MV_ASD$ADOS_CSS_SA.y<-as.numeric(MV_ASD$ADOS_CSS_SA.y)

MV_ASD_Verbal<- MV_ASD[which(MV_ASD$lvis_NVnow==0),]
MV_ASD_MV<- MV_ASD[which(MV_ASD$lvis_NVnow==1),]

summary(MV_ASD_Verbal$ADOS_CSS_SA.x)
sd(MV_ASD_Verbal$ADOS_CSS_SA.x, na.rm = TRUE)

summary(MV_ASD_Verbal$ADOS_CSS_RRB.x)
sd(MV_ASD_Verbal$ADOS_CSS_RRB.x, na.rm = TRUE)

summary(MV_ASD_MV$ADOS_CSS_SA.x)
sd(MV_ASD_MV$ADOS_CSS_SA.x, na.rm = TRUE)

summary(MV_ASD_MV$ADOS_CSS_RRB.x)
sd(MV_ASD_MV$ADOS_CSS_RRB.x, na.rm = TRUE)

```



